---
title: "TreeSE_Coxmodel_3"
format: html
editor: visual
---

## This report focuses on evaluating the predictive performance of the Cox proportional hazards model under transformed and un-transformed data. The primary objective is to identify which transformation method allows the Cox model to yield the most accurate survival predictions. By systematically applying distinct transformations to the data-set and assessing the model’s performance, this study aims to provide insights into optimizing data pre-processing for survival analysis.

Load libraries

```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(mia)
library(ggplot2)
library(lattice)
library(TreeSummarizedExperiment)
library(coda4microbiome)
library(SummarizedExperiment)
library(tibble)
library(phyloseq)
library(survival)
library(survival)
```

link to download the biom.Rdata dataset: https://github.com/wg99526/MiSurvGit/tree/main/Data

```{r}
load("~/Desktop/thesis work/draft_experiments/Microbiome_Analysis/TreeSE_survival_analysis/biom.Rdata")
```

convert phyloseq object to TreeSE

```{r, message=FALSE, warning=FALSE, echo=TRUE}
tse <- convertFromPhyloseq(biom)
```

```{r}
library(mia)

# 1. Agglomerate features at Genus level
tse_genus <- agglomerateByRank(tse, rank = "Genus")

# Check structure
tse_genus

# 2. Filter out taxa that are zero in more than 95% of samples
# Calculate prevalence for each feature
prev <- apply(assay(tse_genus), 1, function(x) sum(x > 0) / length(x))

# Keep taxa present in at least 5% of samples
keep <- prev >= 0.05
tse_filtered <- tse_genus[keep, ]

# Check structure after filtering
tse_filtered

```

prepare transformed assay

```{r}
#all pair-wise log ratio format
coef_threshold = 0
lambda = "lambda.1se" 
nvar = NULL
alpha = 0.9
nfolds = 5
showPlots = TRUE
x1 <- impute_zeros(assay(tse_filtered))
x1 <- t(x1)
kselect <- ncol(x1)
taxaselect <- (1:ncol(x1))
lrmatrix <- logratios_matrix(x1)
x_lrX <- lrmatrix[[1]] #all pair-wise log ratio transformed assay
x_lrX <- t(x_lrX)
idlrX <- lrmatrix[[2]]
```

Prepare the TreeSummarizedExperiment that contain several assays

```{r, message=FALSE, warning=FALSE, echo=TRUE}
tse_filtered <- transformAssay(tse_filtered, method = 'relabundance')
tse_filtered <- transformAssay(tse_filtered, method = 'clr', pseudocount = T)
```

Add alternative experiment to the TreeSE

```{r}
library(SummarizedExperiment)
se <- SummarizedExperiment(assays = list(counts = x_lrX))
altExp(tse_filtered, "all pair-wise log ratio") <- se
altExpNames(tse_filtered)
```

split data

```{r}
set.seed(123)

# Get number of samples
num_samples <- ncol(tse_filtered)

# Split into training and test indices
train_indices <- sample(seq_len(num_samples), size = floor(0.7 * num_samples))
test_indices <- setdiff(seq_len(num_samples), train_indices)

# Subset the TreeSummarizedExperiment object
tse_train <- tse_filtered[, train_indices]
tse_test  <- tse_filtered[, test_indices]
```

prepare training data

```{r}
col_data_train <- as.data.frame(colData(tse_train))
# Convert all assays to matrices
x_counts_train       <- t(as.matrix(assay(tse_train, "counts")))
x_clr_train          <- t(as.matrix(assay(tse_train, "clr")))
x_relabundance_train <- t(as.matrix(assay(tse_train, "relabundance")))
x_aplr_train         <- t(as.matrix(assay(altExp(tse_train, "all pair-wise log ratio")))) 
# Combine all assays and alternativeexperiemnt
x_full_train <- cbind(x_counts_train, x_clr_train, x_relabundance_train, x_aplr_train)
# Replace row names with only digits extracted from original row names
rownames(x_full_train) <- gsub("\\D", "", rownames(x_full_train))
colnames(x_full_train) <- gsub("\\D", "", colnames(x_full_train))
time_train <- col_data_train$T1Dweek   # numeric vector
status_train <- col_data_train$T1D  # binary vector (0 = censored, 1 = event)
covars_train <- col_data_train[, c("Antibiotics", "Sex")]
```

prepare testing data

```{r}
#prepare the testing subset
col_data_test <- as.data.frame(colData(tse_test))
# Extract time-to-event and status from colData
time_test <- col_data_test$T1Dweek   # numeric vector
status_test <- col_data_test$T1D  # binary vector (0 = censored, 1 = event)
covars_test <- col_data_test[, c("Antibiotics", "Sex")]  # update with correct column names
# Extract abundance matrix (assumes 1st assay)
# Convert all assays to matrices
x_counts_test       <- t(as.matrix(assay(tse_test, "counts")))
x_clr_test          <- t(as.matrix(assay(tse_test, "clr")))
x_relabundance_test <- t(as.matrix(assay(tse_test, "relabundance")))
x_aplr_test         <- t(as.matrix(assay(altExp(tse_test, "all pair-wise log ratio"))))
# Combine all assays and alternativeexperiemnt
x_full_test <- cbind(x_counts_test, x_clr_test, x_relabundance_test, x_aplr_test)
# Replace row names with only digits extracted from original row names
rownames(x_full_test) <- gsub("\\D", "", rownames(x_full_test))
colnames(x_full_test) <- gsub("\\D", "", colnames(x_full_test))
```

store the assays

```{r}
# Assay list from your TreeSummarizedExperiment (TSE)
#1-assays from the training subset of tse object
assay_list_train <- list(
  Counts = x_counts_train,
  CLR = x_clr_train,
  RelativeAbundance = x_relabundance_train,
  APLR = x_aplr_train
)
#2-assays from the testing subset of tse object
assay_list_test <- list(
  Counts = x_counts_test,
  CLR = x_clr_test,
  RelativeAbundance = x_relabundance_test,
  APLR = x_aplr_test
)
```

```{r}
set.seed(123)
y_test = Surv(time_test, status_test)# Cox response variable (time & status)
y = Surv(time_train, status_train)# fits a standard Cox proportional hazards model (coxph) using all covariate
df1 <- data.frame(as.matrix(y), covars_train)
model1 <- coxph(Surv(time_train, status_train) ~ ., data = df1)
x1 <- predict(model1)#linear predictor
```

implement cox model by looping over all assays/altExps

```{r}
# 1. Create empty list to store models
cvfit_list <- list()

# 2. First loop: Fit a separate model for each assay
for (assay_name in names(assay_list_train)) {
  x_assay_train <- assay_list_train[[assay_name]]
  
  cvfit <- glmnet::cv.glmnet(x_assay_train, y,
                             family = "cox",
                             type.measure = "C",
                             nfolds = 5,
                             alpha = alpha,
                             keep = TRUE,
                             offset = x1)
  
  # Store the trained model
  cvfit_list[[assay_name]] <- cvfit
}
```

```{r}
# 3. Create a named vector to store C-indices
cindex_results <- numeric(length(assay_list_test))
names(cindex_results) <- names(assay_list_test)

# 4. Second loop: Predict on test set and compute C-index
for (assay_name in names(assay_list_test)) {
  x_assay_test <- assay_list_test[[assay_name]]
  cvfit <- cvfit_list[[assay_name]]  # retrieve matching model
  lambda_best <- cvfit$lambda.min    # or use a fixed lambda if you want
  
  # Predict on test set
  predictions <- as.numeric(predict(cvfit, newx = x_assay_test, s = lambda_best, newoffset = x1))
  
  # Compute and store C-index
  cindex_results[assay_name] <- glmnet::Cindex(pred = predictions, y_test)
}

```

```{r}
print(cindex_results)
```

```{r}
# Number of folds used in cross-validation
nfolds <- 5

# Initialize matrix to store C-indices: rows = assays, cols = folds
cindex_matrix <- matrix(NA, nrow = length(cvfit_list), ncol = nfolds)
rownames(cindex_matrix) <- names(cvfit_list)
colnames(cindex_matrix) <- paste0("Fold", 1:nfolds)

# Compute C-index per fold per assay
for (assay_name in names(cvfit_list)) {
  cvfit <- cvfit_list[[assay_name]]
  
  # Get per-fold predictions from the cross-validation
  fit_preval <- cvfit$fit.preval  # matrix of predictions: rows = samples, cols = lambdas
  foldid <- cvfit$foldid          # which fold each sample was in
  lambda_index <- which(cvfit$lambda == cvfit$lambda.min)
  
  for (k in 1:nfolds) {
    test_indices <- which(foldid == k)
    preds <- fit_preval[test_indices, lambda_index]
    y_fold <- y[test_indices]
    
    # Compute C-index for that fold and assay
    cindex_matrix[assay_name, k] <- glmnet::Cindex(pred = preds, y_fold)
  }
}

# View the C-index matrix
print(cindex_matrix)

```

```{r}
library(reshape2)  # for melting the matrix
library(ggplot2)

# Convert matrix to long format dataframe
df_cindex <- melt(cindex_matrix, varnames = c("Assay", "Fold"), value.name = "Cindex")

# Plot boxplot
ggplot(df_cindex, aes(x = Assay, y = Cindex)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  theme_minimal() +
  labs(title = "C-index Distribution Across Assays",
       x = "Assay Version",
       y = "C-index") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Test the difference between the C-indices

```{r}
kruskal_test_result <- kruskal.test(Cindex ~ Assay, data = df_cindex)
print(kruskal_test_result)
```

```{r}
friedman_result <- friedman.test(as.matrix(cindex_matrix))
# Print the result
print(friedman_result)
```
